---
title: "Bayesian Additive Regression Trees (BART)"
author:
  - "Jake Fortuna"
  - "Tyler Johnson"
  - "Ryan Pini"
  - "Ashley Powell"
  - "Samika Reddy"
  - "Kathleen Reece"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    self_contained: true
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 4.5,
  fig.retina = 3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)


```

---

## Baby Example: BART vs LOESS Regression

We compare two regression methods:

- **LOESS**
- **BART (Bayesian Additive Regression Trees)**

Goal:

- Predict motorcycle acceleration from time
- Compare training and testing error
- Connect BART to a method already used in class

---

## Step 1: Train/Test Split

```{r echo=TRUE, results='hide'}
# Load the packages we need
library(MASS)
install.packages("dbarts")
library(dbarts)

# Set a seed so results are reproducible
set.seed(123)

# Load the mcycle dataset
data(mcycle)

# Rename variables for clarity
x <- mcycle$times
y <- mcycle$accel

# Count the total number of observations
n <- length(y)

# Randomly choose about 70% of the observations for training
train_idx <- sample(1:n, size = floor(0.7 * n))

# Create training and testing sets
x_train <- x[train_idx]
y_train <- y[train_idx]

x_test <- x[-train_idx]
y_test <- y[-train_idx]

# BART needs predictor variables to be matrices
x_train_bart <- as.matrix(x_train)
x_test_bart  <- as.matrix(x_test)
```

---
Step 2: Fit LOESS Models

```{r}
# Define the spans we want to compare
spans <- c(0.25, 0.35, 0.50, 0.65, 0.75)

# Create vectors to store LOESS training and testing MSE
loess_train_mse <- numeric(length(spans))
loess_test_mse  <- numeric(length(spans))

# Loop through each span value
for (i in 1:length(spans)) {
  
  # Fit LOESS using the training data
  loess_fit <- loess(
    y_train ~ x_train,
    span = spans[i],
    degree = 1
  )
  
  # Predict on the training data
  loess_train_pred <- predict(
    loess_fit,
    newdata = data.frame(x_train = x_train)
  )
  
  # Predict on the testing data
  loess_test_pred <- predict(
    loess_fit,
    newdata = data.frame(x_train = x_test)
  )
  
  # Compute training and testing MSE
  loess_train_mse[i] <- mean((y_train - loess_train_pred)^2, na.rm = TRUE)
  loess_test_mse[i]  <- mean((y_test - loess_test_pred)^2, na.rm = TRUE)
}
```


---

Step 3: Fit BART Model
```{r}
# Fit the BART model
bart_model <- wbart(
  x.train = x_train_bart,
  y.train = y_train,
  x.test  = x_test_bart,
  ndpost  = 200
)

# Extract posterior prediction matrices
bart_train_mat <- bart_model$yhat.train
bart_test_mat  <- bart_model$yhat.test

# Make sure rows are posterior draws and columns are observations
if (ncol(bart_train_mat) == length(y_train)) {
  bart_train_draws <- bart_train_mat
} else {
  bart_train_draws <- t(bart_train_mat)
}

if (ncol(bart_test_mat) == length(y_test)) {
  bart_test_draws <- bart_test_mat
} else {
  bart_test_draws <- t(bart_test_mat)
}

# Count the number of posterior draws
n_draws <- nrow(bart_train_draws)

# Create vectors to store BART training and testing MSE
bart_train_mse <- numeric(n_draws)
bart_test_mse  <- numeric(n_draws)

# Compute MSE for each posterior draw
for (i in 1:n_draws) {
  bart_train_mse[i] <- mean((y_train - bart_train_draws[i, ])^2)
  bart_test_mse[i]  <- mean((y_test - bart_test_draws[i, ])^2)
}

# Smooth the BART errors using cumulative averages
bart_train_mse_smooth <- cumsum(bart_train_mse) / seq_along(bart_train_mse)
bart_test_mse_smooth  <- cumsum(bart_test_mse) / seq_along(bart_test_mse)
```
---
Step 4: LOESS Error Comparison
```{r}
plot(spans, loess_train_mse,
     type = "b",
     pch = 19,
     lwd = 2,
     ylim = range(c(loess_train_mse, loess_test_mse), na.rm = TRUE),
     xlab = "LOESS Span",
     ylab = "Mean Squared Error",
     main = "LOESS Train and Test Error")

lines(spans, loess_test_mse,
      type = "b",
      pch = 19,
      lty = 2,
      lwd = 2)

legend("topright",
       legend = c("Train MSE", "Test MSE"),
       lty = c(1, 2),
       pch = c(19, 19),
       lwd = 2)
```
LOESS depends strongly on the span value. Smaller spans allow the curve to follow the training data more closely, which can reduce training error but may increase overfitting. Larger spans produce smoother curves, but if the span is too large, the model may oversmooth and miss important patterns in the data.

---

Step 5: BART Error Comparison

```{r}

plot(bart_train_mse_smooth,
     type = "l",
     lwd = 2,
     ylim = range(c(bart_train_mse_smooth, bart_test_mse_smooth)),
     xlab = "Posterior Draw",
     ylab = "Mean Squared Error",
     main = "BART Train and Test Error")

lines(bart_test_mse_smooth,
      lty = 2,
      lwd = 2)

legend("topright",
       legend = c("Train MSE", "Test MSE"),
       lty = c(1, 2),
       lwd = 2)
```
BART produces predictions by averaging over many regression trees. Instead of choosing one final tree, BART uses many posterior draws. The training and testing errors stabilize as more posterior draws are averaged. This shows how BART uses Bayesian regularization to avoid fitting the training data too aggressively.

---
Step 6: Final Comparison
```{r}
# Best LOESS test MSE
best_loess_index <- which.min(loess_test_mse)

comparison_table <- data.frame(
  Method = c("LOESS", "BART"),
  Tuning_or_Draw = c(
    paste("Best span =", spans[best_loess_index]),
    paste("Posterior draws =", n_draws)
  ),
  Train_MSE = c(
    loess_train_mse[best_loess_index],
    mean(bart_train_mse)
  ),
  Test_MSE = c(
    loess_test_mse[best_loess_index],
    mean(bart_test_mse)
  )
)

comparison_table

```


---
The LOESS model requires us to choose a span value, and the final performance depends on that choice. This connects directly to the class discussion of smoothing and the bias-variance tradeoff. A smaller span can have lower bias but higher variance, while a larger span can have higher bias but lower variance.

BART also uses trees, but unlike a single tree, it combines many trees through a Bayesian framework. This regularization helps BART avoid overfitting while still capturing nonlinear patterns in the motorcycle acceleration data.

Overall, this baby example shows that BART can be used for regression problems like the mcycle dataset, and it provides a useful comparison to LOESS, which was already discussed in class.

Main Takeaways
LOESS is useful for smooth nonlinear regression.
LOESS performance depends on the span.
Small spans can overfit.
Large spans can oversmooth.
BART uses many trees and Bayesian averaging.
BART can capture nonlinear patterns without manually choosing a smoothing span.
This example connects BART to a regression method already covered in class.
